---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# wheredd

<!-- badges: start -->
[![R-CMD-check](https://github.com/belian-earth/wheredd/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/belian-earth/wheredd/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/belian-earth/wheredd/graph/badge.svg)](https://app.codecov.io/gh/belian-earth/wheredd)
[![License:Apache](https://img.shields.io/github/license/Permian-Global-Research/vrtility)](https://www.apache.org/licenses/LICENSE-2.0) 
[![CRANstatus](https://www.r-pkg.org/badges/version/vrtility)](https://cran.r-project.org/package=vrtility)

<!-- badges: end -->

**wheredd** provides streamlined access to forest carbon offset project boundary data. The package builds a local DuckDB database from the [CECIL forest carbon boundaries dataset](https://source.coop/cecil/forest-carbon-boundaries), making it easy to query and analyze REDD+ and carbon offset project locations.

**This package focuses on data access.** Once you have the data, use standard R spatial and data analysis tools (sf, dplyr, terra, etc.) for your analysis workflows.

## Installation

Install the development version from GitHub:

```r
# install.packages("pak")
pak::pak("belian-earth/wheredd")
```

## Quick Start

Build the database (downloads pre-processed data from GitHub releases):

```{r quickstart}
library(wheredd)

# just get the urls and download yourself...
carbon_proj_release_url()
# or for access to the original source data urls...
carbon_proj_source_urls()


# Create local database
db_path <- carbon_proj_db()

```

## Usage

### Connect and query

Use standard DBI/duckdb/duckplyr/dplyr workflows:

```{r usage, message=FALSE}
library(DBI)
library(dplyr)
library(duckplyr)
# Connect to database
con <- dbConnect(duckdb::duckdb(), dbdir = db_path, read_only = TRUE)

# Query with SQL
projects <- dbGetQuery(con, "
  SELECT * FROM carbon_projects
  WHERE continent = 'africa' AND area_role = 'project'
  LIMIT 10
")

# Or use d(uck)plyr
projects <- tbl(con, "carbon_projects") |>
  filter(continent == "africa", area_role == "project") |>
  collect()

projects

dbDisconnect(con, shutdown = TRUE)
```

### Work with spatial data

Convert WKB geometries to sf objects:

```{r spatial}
library(sf)

# Read WKB geometry column as sf
projects_sf <- projects |>
  st_as_sf(crs = 4326)

# Now use standard sf operations
plot(projects_sf['id'], axes=TRUE)
```

### Build from source

For fresh data with full geometry processing:

```{r source, eval=FALSE}
# Build from source parquet files (slower, but fresh)
db_path <- carbon_proj_db(
  build_from = "source",
  continents = c("africa", "asia")
)
```

## Data Structure

The database contains one table (`carbon_projects`) with:

- **One row per project per area type** (project, accounting, reference)
- **area_role**: Type of boundary (project/accounting/reference)
- **geometry_wkb**: Geometry as WKB BLOB (use `sf::st_as_sf(wkb = ...)`)
- **Metadata**: Project name, registry, dates, country, etc.
- **Ordered by**: continent → country → id

## Data Source

Data is from the CECIL project's forest carbon boundaries dataset:
https://source.coop/cecil/forest-carbon-boundaries

The package does not include analytical functions—it provides clean, queryable access to the data. Use R's rich ecosystem of spatial and analytical packages for your analysis needs.

## Data Dictionary

The `carbon_projects` table contains the following columns:

| Column | Type | Description |
|--------|------|-------------|
| `id` | VARCHAR | Unique project identifier combining registry abbreviation and project number (e.g., "VCS1234") |
| `project_name` | VARCHAR | Name of the carbon project as documented by the registry |
| `area_role` | VARCHAR | Type of boundary geometry: "project" (implementation area), "accounting" (area for credit calculation), or "reference" (area for baseline trends) |
| `registry_name` | VARCHAR | Carbon registry hosting the project (American Carbon Registry, BioCarbon Registry, Climate Action Reserve, EcoRegistry, Gold Standard, Verra) |
| `methodology` | VARCHAR | Methodology used for project implementation (e.g., VM0015, ACR Methodology) |
| `project_type` | VARCHAR | Type of forestry carbon offset program: ARR (Afforestation/Reforestation), AD (Avoided Deforestation), IFM (Improved Forest Management) |
| `continent` | VARCHAR | Continent where the project is located (africa, asia, europe, north_america, oceania, south_america) |
| `country` | VARCHAR | Country where the project is located |
| `project_developer` | VARCHAR | Entity or individual organizing the carbon offset project |
| `project_start_date` | DATE | Start date of the crediting period |
| `project_end_date` | DATE | End date of the crediting period |
| `entry_date` | DATE | Date when project information was added to the source database |
| `processing_approach` | VARCHAR | Method used to obtain boundary data: "Official" (from project developer), "Georeferenced" (from documents), "Linear" (traced from maps), or "Method" (derived from methodology) |
| `pd_declined` | VARCHAR | Whether the project developer declined to provide geometry (Yes/No/N/A) |
| `filename` | VARCHAR | Source parquet file URL from which this record was read |
| `geometry_wkb` | BLOB | Well-Known Binary representation of the project boundary (convert to sf using `st_as_sf(wkb = "geometry_wkb", crs = 4326)`) |

```{=gfm}
> [!NOTE]
> - Each project may appear 1-3 times depending on available boundary types (project/accounting/reference areas)
> - Empty geometries are filtered out during database creation
> - All geometries are validated, forced to 2D (Z coordinates removed), and stored as WKB BLOBs
> - Both point and polygon geometries are supported (collections are extracted to individual types)
> - Data is ordered by continent → country → id.
``` 

## Original Data Info and Attributions:
**source**: https://source.coop/cecil/forest-carbon-boundaries

**Converted by:** Alex Logan

**Original Authors:** Akshata Karnik, Jack B. Kilbride, Tristan R.H. Goodbody, Rachael Ross, Elias Ayrey (Corresponding Author)

**Date:** June 5, 2024

**License:** CC-BY 4.0 See https://zenodo.org/records/11459391